\chapter{Vector Similarity Search with FAISS}
\label{chapter:similarity}

The system uses vector similarity search to identify semantically related books. 
Once the text embeddings are generated, the task becomes finding the most similar vectors to a given query vector. 
This is accomplished using FAISS — Facebook AI Similarity Search\cite{faiss}.

\section{Similarity Search Theory}
\label{sec:similarity-theory}

Given a user query $q$ and a set of embedded book descriptions $\{\vec{b}_1, \vec{b}_2, \dots, \vec{b}_n\}$, the system computes similarity using the L2 distance:

\[
\text{dist}(\vec{q}, \vec{b}_i) = \|\vec{q} - \vec{b}_i\|_2^2
\]

Where:
\begin{itemize}
    \item $\vec{q}$ is the embedding of the user query.
    \item $\vec{b}_i$ is the embedding of the $i$th book description.
    \item The top $k$ results with the smallest distance values are returned as recommendations.
\end{itemize}

This approach relies on the assumption that semantically similar texts lie close together in the embedding space.

---

\section{Why FAISS?}
\label{sec:faiss-choice}

FAISS is a library optimized for fast similarity search of dense vectors. It is designed for high-dimensional, large-scale applications and supports both exact and approximate nearest neighbor search.

\begin{itemize}
    \item \textbf{Speed:} Fast search even for large datasets.
    \item \textbf{Flexibility:} Multiple indexing structures and distance metrics.
    \item \textbf{Local Execution:} FAISS supports CPU-only environments and offline usage.
\end{itemize}

In this project, FAISS was used in its simplest form: \texttt{IndexFlatL2}, which performs exact nearest-neighbor search using Euclidean distance.

---

\section{Implementation}
\label{sec:similarity-implementation}

After generating the embeddings, the FAISS pipeline performs the following steps:

\begin{enumerate}
    \item Create an index using \texttt{faiss.IndexFlatL2(dimension)}.
    \item Add the matrix of 6,507 book vectors to the index.
    \item Save the index to disk as \texttt{index.faiss} in the \texttt{embeddings/} folder.
    \item At query time:
    \begin{itemize}
        \item The user query is embedded using the same MiniLM model.
        \item The query vector is compared against the index to find the top-$k$ nearest neighbors.
        \item Metadata for matching books is retrieved from \texttt{books\_indexed.csv}.
    \end{itemize}
\end{enumerate}

---

\section{Summary}
\label{sec:similarity-summary}

FAISS allows the system to perform efficient and scalable semantic similarity search over thousands of books. 
By combining powerful sentence embeddings with fast vector indexing, the recommender system delivers meaningful results in real time—entirely offline.
