\chapter{User Interface}
\label{chapter:interface}

The user interface enables interaction with the semantic book recommendation system. 
While not a machine learning component, it plays a crucial role in usability, filtering, and presentation. 
The interface is implemented using \texttt{Streamlit}, a lightweight Python framework suitable for local-first applications.

\section{Design Principles}
\label{sec:ui-design}

The UI was designed with simplicity and responsiveness in mind. It supports end-to-end offline operation. Design goals included:

\begin{itemize}
    \item Concept-based natural language search
    \item Multi-faceted filtering by category and rating
    \item Minimal dependencies and fast startup
    \item Clear presentation of recommendations
\end{itemize}

All inference and retrieval logic runs locally.

\section{Query Workflow}
\label{sec:ui-query-workflow}

The system accepts a free-form text query from the user (e.g., \textit{``Books about surviving on Mars''}). The workflow proceeds as follows:

\begin{enumerate}
    \item Query is embedded using the MiniLM model.
    \item The FAISS index returns the top 60 nearest book vectors.
    \item Filtering and sorting options are applied.
    \item Six book cards are rendered per page.
\end{enumerate}

This approach allows expressive, language-based search.

\section{Filtering and Sorting}
\label{sec:ui-filtering}

Users can refine results with:

\begin{itemize}
    \item \textbf{Category filter:} Multi-select widget for genre labels
    \item \textbf{Rating sort:} Toggle for ascending/descending order
    \item \textbf{Pagination:} Adjustable page selection for navigation
\end{itemize}

Filters are applied after FAISS similarity search.

\section{Result Display}
\label{sec:ui-display}

Each result is rendered as a card containing:

\begin{itemize}
    \item Book cover image (URL with fallback)
    \item Full title and author
    \item Average rating, number of pages, and publication year
    \item Description excerpt (up to 300 characters)
    \item Refined categories
\end{itemize}

The layout ensures readability and intuitive navigation.

\section{Privacy and Offline Execution}
\label{sec:ui-privacy}

The application is designed for full offline use:

\begin{itemize}
    \item No tracking, cookies, or login
    \item No data transmission to external servers
    \item Book data, models, and index are stored locally
\end{itemize}

This ensures data privacy, fast load times, and compatibility with low-resource environments.

\section{Summary}
\label{sec:ui-summary}

The Streamlit interface allows users to interact with the system seamlessly. 
Natural language search, combined with real-time filtering and semantic recommendations, creates an intuitive and private discovery experience.